Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-07-20 paper.tex
Expand homogeneous equilibrium text

Greek tensor indices ($\alpha, \beta, ...$) range over all four coordinates,
whereas Latin indices ($i, j, ...$) range over the spatial coordinates 1--3.
We use naturalized units in which $\hbar=c=1$.
And for most of the remainder of this article we suppress neutrino species label
$\nu_\sigma$ since the formulation is general to any species.
Where we do reference particular species we use the three categories
$\nu_e$, $\bar{\nu}_e$, and
$\nu_x=\left\{\nu_\mu,\bar{\nu}_\mu,\nu_\tau,\bar{\nu}_\tau\right\}$.

@@

In the following we present two forms of the integrated distribution functions,
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse}:
1) the full form  by which the solution is $f_{\rm scat}=f_{\rm AE}+f_{\rm SE}$
(labeled \emph{scat} since we include elastic scattering), and
2) the form treating only absorption/emission interactions
in which we set $\mathscr{K}_{\rm SE}=0$ so that $f_{\rm noscat}=f_{\rm AE}$
(labeled \emph{noscat} since we ignore elastic scattering).
There is no simple relation between the two forms for a given ray; that is
$f_{\rm scat} \neq f_{\rm noscat}+f_{\rm SE}$.
This is because $f_{\rm AE}$ is integrated using solely the absorption optical
depth in the first form,
and the absorption plus scattering optical depth in the second form.
The differences between these two methods is apparent
in Fig.~\ref{fig:cumulative_f_homogeneous}.
Before this work, the \emph{noscat} form of the equations has been the standard for
\todo{explain why; cite others}
ray tracing, though many authors include the scattering optical depth in
the integration of Eqn.~\ref{eqn:num_fae}
(see for example \cite{hari2010-gr_nunubar_collapsar}).

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at the radius 50~km.
For comparision with that test we use the LS180 equation of state
\cite{latt1991-nuc_eos}
\footnote{available at \lstinline{stellarcollapse.org}}
in which the equilibrium $\nu_e$ neutrino chemical potential is
$\mu_{\nu_e}=-0.575$~MeV
(with $\mu_{\bar{\nu}_e}=-\mu_{\nu_e}$ and $\mu_{\nu_x}=0$).
Fig.~\ref{fig:mfps-50km} presents the neutrino mean free paths
at this thermodynamic point over two energy decades.
For this test we set up our slab large enough for neutrinos of all
energies to be trapped. 
Since our ray tracing algorithm integrates rays to terminal optical
depths of $\tau_{\rm term}>14$, our domain must extend to $r=10^7\,{\rm km}$.

Fig.~\ref{fig:cumulative_f_homogeneous} displays the cumulative distribution
function integrated along the ray, for each of the three species and using
both methods, \emph{noscat} and \emph{scat}. We display
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse} in their integral form:
\begin{equation}
  \label{eqn:cumulative_f}
  f(t)
  = - \int_{0}^t dt'\,\frac{1}{p^t}e^{-\tau(0,t')}
  \mathscr{K}(t')f^{\rm eq}(t'),
\end{equation}
and the integration proceeds backwards in time, from $t=0$
\todo{could write more simply because Minkowski}
to some terminal $t<0$.
The figure shows this backwards-in-time integration proceeding left to right.

In this test and only this test, we used a higher-order integrator, a 5th
order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed}. This is because
\todo{discuss with Matt}
our default 3rd order Runge-Kutta algorithm estimates a vanishing error
in this configuration.

Fig.~\ref{fig:homogeneous_isotropic} shows the error in our results.
\todo{describe with scat on}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-cumulative_f-homogeneous}
  \caption{Cumulative distribution functions at $\varepsilon=11.25$~MeV
    in homogeneous infinite slab of Sec.~\ref{ssec:test_equilibrium}.
    The integration proceeds from left to right (backwards in time $t$).
    The two components of the \emph{scat} solutions ($f_{\rm AE}$ in points,
    and $f_{\rm SE}$ in dashed lines) sum to equal the single component
    of the \emph{noscat} solution ($f_{\rm AE}$ in points connected by lines).
    All solutions terminate when they achieve a total optical depth
    greater than $\tau_term=14$; in the \emph{scat} cases this occurs at an
    earlier time since the total mean free paths are less than the
    absorption mean free paths (much less in the case of $\nu_x$).
    The positions of the points (only plotted for the $f_{\rm AE}$ components)
    are the time steps chosen by the adaptive time-stepping algorithm
    described in Sec.~\ref{ssec:numerical}.
  }
  \label{fig:cumulative_f_homogeneous}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figure-ferr-50km}
  \caption{Relative error in integrated equilibrium distribution functions
    in the infinite homogeneous slab  described in
    Sec.~\ref{ssec:test_equilibrium}.
    Presented here are the relative differences between the final
    $f_{\rm AE}$ in the \emph{noscat} case
    (see Fig.~\ref{fig:cumulative_f_homogeneous})
    and the equilibrium distribution functions given by
    Eqn.~\ref{eqn:feq}.
    As can be seen in Fig.~\ref{fig:mfps-50km} these three species
    over the given energies
    probe our numerical solution over length scales from
    100~m to $10^{5}\,{\rm km}$.
    The source of this error is the discarded boundary term
    $f_{\rm bdry}\sim10^{-6}$, described in Sec.~\ref{ssec:numerical}.
  }
  \label{fig:homogeneous_isotropic}
\end{figure}

@@

\subsection{Post-Bounce Collapse Profile:
  Testing Scattering}
\label{ssec:test_collapse}
[Standard radiation transport test described in~\cite{ocon2015-gr1d_with_nu},
and in \cite[App.~E.6]{fouc2015-m1_nsbh},
and in~\cite{abdi2012-monte_carlo}.]
\todo{clarify inelastic scattering and velocity terms}
Plot in Fig.~\ref{fig:avg_eps_collapse}.

@@

However moment evolutions with multiple energy groups are still rare.
Most simulations employ a gray moment scheme,
evolving only the energy-integrated moments, $J$ and $H^\alpha$
and sometimes the number density $G$.
If such is the case we resort to the \emph{gray} method,
by reconstructing the energy-dependent source terms
from gray moments, assuming a diluted Fermi-Dirac spectrum.
We use the following procedure:
\begin{enumerate}
\item
  Interpolate the fluid temperature, $T$,
  equilibrium neutrino chemical potential, $\mu_\nu$,
  and evolved neutrino energy and number densities in the fluid frame,
  $J$ and $G$, from the simulation grid.
\item
  Compute the average neutrino energy in the fluid frame
  \begin{equation}
    \langle \varepsilon \rangle \equiv J/G.
  \end{equation}
\item
  Compute the neutrino spectral temperature, assuming equilibrium with the
  fluid
  \begin{equation}
    T_\nu = \langle \varepsilon \rangle
    \frac{\mathscr{F}_2(\mu_\nu/T)}{\mathscr{F}_3(\mu_\nu/T)},
  \end{equation}
  where $\mathscr{F}_b$ is the Fermi integral
  \begin{equation}
    \mathscr{F}_b(\eta) = \int_0^\infty dx \, x^b (1+e^{x-\eta})^{-1}.
  \end{equation}
\item
  Assume the background neutrino fields have the same total density
  as the evolved moments
  \begin{align}
    \label{eqn:J_from_gray}
    J(\varepsilon) &=
    J \frac{\varepsilon^3}{T_\nu^4 \,\mathscr{F}_3(\mu_\nu/T)}
    (1+\exp(\varepsilon/T_\nu-\mu_\nu/T)^{-1}, \\
    \label{eqn:H_from_gray}
    H^\mu(\varepsilon) &=
    H^\mu \frac{\varepsilon^3}{T_\nu^4\, \mathscr{F}_3(\mu_\nu/T)}
    (1+\exp(\varepsilon/T_\nu-\mu_\nu/T)^{-1}.
  \end{align}
\end{enumerate}
When we use these moments in Eqn.~\ref{eqn:background_phi},
we call this the \emph{gray} method.

>>>>

Greek tensor indices ($\alpha, \beta, ...$) range over all four coordinates,
whereas Latin indices ($i, j, ...$) range over the spatial coordinates 1--3.
We use naturalized units in which $\{\hbar,c,k_{\rm B}\}=1$.
And for most of the remainder of this article we suppress neutrino species label
$\nu_\sigma$ since the formulation is general to any species.
Where we do reference particular species we use the three categories
$\nu_e$, $\bar{\nu}_e$, and
$\nu_x=\left\{\nu_\mu,\bar{\nu}_\mu,\nu_\tau,\bar{\nu}_\tau\right\}$.

@@

In the following we present two forms of the integrated distribution functions,
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse}:
1) the full form  by which the solution is $f_{\rm scat}=f_{\rm AE,scat}+f_{\rm SE,scat}$
(labeled \emph{scat} since we include elastic scattering), and
2) the form treating only absorption/emission interactions
in which we set $\mathscr{K}_{\rm SE}=0$ so that $f_{\rm noscat}=f_{\rm AE,noscat}$
(labeled \emph{noscat} since we ignore elastic scattering).
There is no simple relation between these forms
(i.e.\ $f_{\rm AE,scat} \neq f_{\rm AE,noscat}$)
because $f_{\rm AE}$ is integrated using the absorption plus scattering optical depth
in the first form and solely the absorption optical depth in the second form.
The differences between the two methods is apparent
in Fig.~\ref{fig:cumulative_f_homogeneous}.
Before this work, the \emph{noscat} form of the equations has been the standard for
\todo{explain why; cite others}
ray tracing, though many authors include the scattering optical depth in
the integration of Eqn.~\ref{eqn:num_fae}
(see for example \cite{hari2010-gr_nunubar_collapsar}).

\subsection{Infinite Homogeneous Slab:
  Testing Thermodynamic Equilibrium}
\label{ssec:test_equilibrium}
In an optically thick region the radiation field is in thermodynamic equilibrium
with the matter.
We set up a large slab of matter
representative of the fluid in the Massive Collapsed Star test
(Sec.~\ref{ssec:test_collapse}) at the radius 50~km.
For comparision with that test we use the LS180 equation of state
\cite{latt1991-nuc_eos}
\footnote{available at \lstinline{stellarcollapse.org}}
in which the equilibrium $\nu_e$ neutrino chemical potential is
$\eta_{\nu_e}=-0.1555$
(with $\eta_{\bar{\nu}_e}=-\eta_{\nu_e}$ and $\eta_{\nu_x}=0$).

We use the equilibrium distribution functions to define background
fields for the \emph{scat} case:
in the \emph{spectral} method we use
$J(\varepsilon)=\varepsilon^3 f^{\rm eq}(\eta_\nu,T;\varepsilon)/(2\pi^2)$,
and in the \emph{gray} method we use
$J=c T^4 \mathscr{F}_3(\eta_\nu)/(2\pi^2)$ and
$G=c T^3 \mathscr{F}_2(\eta_\nu)/(2\pi^2)$,
using the Fermi integrals defined in Eqn.~\ref{eqn:fermi_integral},
and with $c$ the conversion constant from ${\rm energy}^3$ to ${\rm length}^{-3}$.

Fig.~\ref{fig:mfps-50km} presents the neutrino mean free paths
at this thermodynamic point over two energy decades.
For this test we set up our slab large enough for neutrinos of all
energies to be trapped.
Our domain extends to $\tau_{\rm term}\,L_{\rm mfp,max}\approx10^7\,{\rm km}$,
since our ray tracing algorithm integrates rays to terminal optical
depths of $\tau_{\rm term}\geq14$.

@@

Fig.~\ref{fig:cumulative_f_homogeneous} displays the cumulative distribution
function integrated along the ray, for each of the three species and using
both methods, \emph{noscat} and \emph{scat}. We display
Eqns.~\ref{eqn:num_fae} and \ref{eqn:num_fse} in their integral form:
\begin{equation}
  \label{eqn:cumulative_f}
  f(t)
  = - \int_{0}^t \frac{dt'}{p^t}e^{-\tau(0,t')}
  \mathscr{K}(t')f^{\rm eq}(t'),
\end{equation}
and the integration proceeds backwards in time, from $t=0$
\todo{could write more simply because Minkowski}
to some terminal $t<0$.
The figure shows this backwards-in-time integration proceeding left to right.

\begin{figure}
  \includegraphics[width=\columnwidth]{fig-cumulative_f-homogeneous}
  \caption{Cumulative distribution functions at $\varepsilon=11.25$~MeV
    in homogeneous infinite slab of Sec.~\ref{ssec:test_equilibrium}.
    The integration proceeds from left to right, or backwards in time $t$.
    Each ray terminates when it achieves a total optical depth
    greater than $\tau_{\rm term}=14$.
    The positions of the points (only plotted for the $f_{\rm AE}$ components)
    are the time steps chosen by the adaptive time-stepping algorithm
    described in Sec.~\ref{ssec:numerical}.
  }
  \label{fig:cumulative_f_homogeneous}
\end{figure}

In Fig.~\ref{fig:cumulative_f_homogeneous} we see some expected features.
The final distribution functions asymptote to their equilibrium Fermi-Dirac
levels $f^{\rm eq}_{\{\nu_e,\bar{\nu}_e,\nu_x\}}=\{0.039, 0.053, 0.046\}$
within a few mean free paths;
in the \emph{scat} cases it is the sums $f_{\rm AE}+f_{\rm SE}$ that
achieve these values.
And the lengths of the rays are proportional to the mean free paths:
$L_{\nu_x}>L_{\bar{\nu}_e}>L_{\nu_e}$;
in the \emph{scat} cases these lengths are less than in the \emph{noscat} cases,
since the total mean free paths are less than the absorption mean free paths
(significantly so for $\nu_x$, negligibly so for $\nu_e$).

As Fig.~\ref{fig:mfps-50km} shows, integrating the distribution functions
over the energies $\varepsilon\in[0,100]$~MeV
probes our numerical solution over length scales from
0.1~km to $10^{5}\,{\rm km}$.
Fig.~\ref{fig:homogeneous_isotropic} shows the error in our results for
the \emph{noscat} case.
In this simple case, the dominant source of error is the neglected
boundary term with a relative scale of $e^{-\tau_{\rm term}}\approx10^{-6}$.
In the following inhomogeneous configurations,
errors from the integration dominate over the boundary term.
In this test and only this test, we used a higher-order integrator, a 5th
order Dormand-Prince algorithm \cite{pres2007-nr_3rd_ed}. This is because
\todo{discuss with Matt}
our default 3rd order Runge-Kutta algorithm estimates a vanishing error
in this configuration.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure-ferr-50km}
  \caption{Relative error in integrated equilibrium distribution functions
    in the infinite homogeneous slab  described in
    Sec.~\ref{ssec:test_equilibrium}.
    Presented here are the relative differences between the final
    $f_{\rm AE}$ in the \emph{noscat} case
    (see Fig.~\ref{fig:cumulative_f_homogeneous})
    and the equilibrium distribution functions given by
    Eqn.~\ref{eqn:feq}.
    The source of this error is the discarded boundary term
    $f_{\rm bdry}\sim10^{-6}$, described in Sec.~\ref{ssec:numerical}.
  }
  \label{fig:homogeneous_isotropic}
\end{figure}

@@

\subsection{Post-Bounce Collapse Profile:
  Testing Scattering}
\label{ssec:test_collapse}
We use a single time snapshot of a supernova model in its accretion phase,
100ms after core bounce.
[Standard radiation transport test described in~\cite{ocon2015-gr1d_with_nu},
and in \cite[App.~E.6]{fouc2015-m1_nsbh},
and in~\cite{abdi2012-monte_carlo}.]
\todo{clarify inelastic scattering and velocity terms}
Plot in Fig.~\ref{fig:avg_eps_collapse}.

@@

However moment evolutions with multiple energy groups are still rare.
Most simulations employ a gray moment scheme,
evolving only the energy-integrated moments, $J$ and $H^\alpha$
and sometimes the number density $G$.
If such is the case we resort to the \emph{gray} method,
by reconstructing the energy-dependent source terms
from gray moments, assuming a diluted Fermi-Dirac spectrum.
We use the following procedure:
\begin{enumerate}
\item
  Interpolate the fluid temperature, $T$,
  equilibrium neutrino chemical potential, $\eta_\nu$,
  and evolved neutrino energy and number densities in the fluid frame,
  $J$ and $G$, from the simulation grid.
\item
  Compute the average neutrino energy in the fluid frame
  \begin{equation}
    \langle \varepsilon \rangle \equiv J/G.
  \end{equation}
\item
  Compute the neutrino spectral temperature, assuming equilibrium with the
  fluid
  \begin{equation}
    T_\nu = \langle \varepsilon \rangle
    \frac{\mathscr{F}_2(\eta_\nu)}{\mathscr{F}_3(\eta_\nu)},
  \end{equation}
  where $\mathscr{F}_b$ is the Fermi integral
  \begin{equation}
    \label{eqn:fermi_integral}
    \mathscr{F}_b(\eta) = \int_0^\infty dx \, x^b (1+e^{x-\eta})^{-1}.
  \end{equation}
\item
  Assume the background neutrino fields have the same total density
  as the evolved moments
  \begin{align}
    \label{eqn:J_from_gray}
    J(\varepsilon) &=
    J \frac{\varepsilon^3}{T_\nu^4 \,\mathscr{F}_3(\eta_\nu)}
    (1+\exp(\varepsilon/T_\nu-\eta_\nu)^{-1}, \\
    \label{eqn:H_from_gray}
    H^\mu(\varepsilon) &=
    H^\mu \frac{\varepsilon^3}{T_\nu^4\, \mathscr{F}_3(\eta_\nu)}
    (1+\exp(\varepsilon/T_\nu-\eta_\nu)^{-1}.
  \end{align}
\end{enumerate}
When we use these moments in Eqn.~\ref{eqn:background_phi},
we call this the \emph{gray} method.
